Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19

### 2024-05-16 manuscript/appendix.qmd
Add Audrey edits

---
title: 'Online appendix for "Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19"'
short-title: 'Online appendix for "Pandemic Pass?"'
subtitle: "(Research note)"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  orcid: 0000-0003-3506-7590
  url: https://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
- name: Audrey L. Comstock
  affiliation: Arizona State University
  orcid: 0000-0003-1772-5350
  url: https://audreylcomstock.weebly.com/
  email: audrey.comstock@asu.edu
- name: Andrew Heiss
  affiliation: Georgia State University
  orcid: 0000-0002-3948-3914
  url: https://www.andrewheiss.com
  email: aheiss@gsu.edu
date: April 30, 2024
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/mountainous-mackerel>"
bibliography: references.json
csl: csl/chicago-author-date.csl
link-citations: true
crossref: 
  tbl-title: "Table&nbsp;A"
  fig-title: "Figure&nbsp;A"
---

@@

options(digits = 3, width = 120,
        dplyr.summarise.inform = FALSE,
        knitr.kable.NA = "")
```

@@

```{r libraries-data}
library(tidyverse)
library(patchwork)
library(modelsummary)
library(scales)
library(kableExtra)
library(gt)
library(tinytable)
library(targets)

@@

>>>>

---
title: 'Online appendix for "Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19"'
short-title: 'Online appendix for "Pandemic Pass?"'
subtitle: "(Research note)"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  orcid: 0000-0003-3506-7590
  url: https://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
- name: Audrey L. Comstock
  affiliation: Arizona State University
  orcid: 0000-0003-1772-5350
  url: https://audreylcomstock.weebly.com/
  email: audrey.comstock@asu.edu
- name: Andrew Heiss
  affiliation: Georgia State University
  orcid: 0000-0002-3948-3914
  url: https://www.andrewheiss.com
  email: aheiss@gsu.edu
date: May 16, 2024
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/mountainous-mackerel>"
bibliography: references.json
csl: csl/chicago-author-date.csl
link-citations: true
crossref: 
  tbl-title: "Table&nbsp;A"
  fig-title: "Figure&nbsp;A"
---

@@

options(digits = 3, width = 120,
        dplyr.summarise.inform = FALSE)
```

@@

```{r libraries-data}
library(tidyverse)
library(patchwork)
library(modelsummary)
library(scales)
library(tinytable)
library(targets)

@@

```{r tbl-achr-echr-iccpr}
#| tbl-cap: States that derogated from regional human rights conventions
#| include: true
#| echo: false

achr_echr_iccpr <- tribble(
                 ~ACHR, ~Also,             ~ECHR, ~Also1,
           "Argentina", "Yes",         "Albania",   "No",
             "Bolivia",  "No",         "Armenia",  "Yes",
               "Chile", "Yes",      "Azerbaijan",  "Yes",
            "Colombia", "Yes",         "Estonia",  "Yes",
  "Dominican Republic", "Yes",         "Georgia",  "Yes",
             "Ecuador", "Yes",          "Latvia",  "Yes",
         "El Salvador", "Yes",         "Moldova",  "Yes",
           "Guatemala", "Yes", "North Macedonia",   "No",
            "Honduras",  "No",         "Romania",  "Yes",
             "Jamaica",  "No",      "San Marino",  "Yes",
              "Panama",  "No",          "Serbia",   "No",
            "Paraguay", "Yes",                NA,     NA,
                "Peru", "Yes",                NA,     NA,
            "Suriname",  "No",                NA,     NA,
           "Venezuela",  "No",                NA,     NA
  )

achr_echr_iccpr %>% 
  as.data.frame() %>% 
  set_names(
    c(
      "States that derogated from the ACHR",
      "Also submitted ICCPR derogation",
      "States that derogated from the ECHR",
      "Also submitted ICCPR derogation"
    )
  ) %>% 
  tt(
    width = c(0.3, 0.2, 0.3, 0.2),
    notes = "ACHR = American Convention on Human Rights; ECHR = European Convention on Human Rights"
  ) %>% 
  format_tt(replace = list(" " = NA)) %>% 
  style_tt(j = 1:4, align = "lclc") %>% 
  style_tt(
    bootstrap_class = "table table-sm",
    tabularray_inner = "cells={font=\\scriptsize}, row{1}={valign=b}"
  )
```

```{r tbl-results-determinants}
#| tbl-cap: Results from Authors (2024) showing the determinants of derogation
#| include: true
#| echo: false

det_models <- readRDS(here::here("data", "raw_data",  "determinants_models.rds"))

modelsummary(
  det_models$models, 
  exponentiate = TRUE,
  estimate = "{estimate} ({std.error}){stars}",
  statistic = NULL,
  coef_map = det_models$coef_map,
  gof_map = det_models$gof_map,
  output = "tinytable",
  fmt = fmt_significant(2),
  stars = c("*" = 0.1, "**" = 0.05, "***" = 0.01),
  notes = list(
    "Dependent variable = derogation submission",
    "Logistic regression models; odds ratios (standard error)",
    "* p < 0.1, ** p < 0.05, *** p < 0.01"
  ),
  width = c(0.5, 0.25, 0.25)
) %>% 
  style_tt(
    bootstrap_class = "table table-sm",
    tabularray_inner = "cells={font=\\scriptsize}"
  )
```

### 2024-05-16 manuscript/manuscript.qmd
Add Audrey edits

---
title: "Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19"
short-title: "Pandemic Pass?"
subtitle: "(Research note)"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  orcid: 0000-0003-3506-7590
  url: https://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
- name: Audrey L. Comstock
  affiliation: Arizona State University
  orcid: 0000-0003-1772-5350
  url: https://audreylcomstock.weebly.com/
  email: audrey.comstock@asu.edu
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  orcid: 0000-0002-3948-3914
  url: https://www.andrewheiss.com
  email: aheiss@gsu.edu
  attributes:
    corresponding: true
correspondence-prefix: "Correspondence concerning this article should be addressed to"
abstract: |
  This research note asks whether states issuing pandemic-era human rights treaty derogations implemented emergency provisions as intended or used them to abuse human rights during a time of crisis. In an effort to combat the COVID-19 pandemic, many countries declared states of emergency and derogated (temporarily suspended) from their international human rights treaty obligations. Using data from the Varieties of Democracy PanDem dataset and the Oxford COVID-19 Government Response Tracker, we find that states that derogated from their international human rights obligations imposed emergency measures that were temporary and did not violate non-derogable rights. On the other hand, states that did not derogate were more likely impose discriminatory measures, enact emergency measures without time limits and violate non-derogable rights. Our results support the role that flexibility mechanisms such as derogations play in international law and show that states are being sincere about their intentions and not, generally, using these mechanisms to cover abusive behavior.
thanks: |
  We would like to thank particpants of the 2022 Peace Science Society conference and the 2023 International Studies Association (ISA) convention for their comments and feedback.
date: April 30, 2024
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/mountainous-mackerel>"
keywords:
  - human rights
  - treaty derogations
  - COVID-19
  - emergency policies
bibliography: references.json
csl: csl/chicago-author-date.csl
link-citations: true
---

@@

The ICCPR is a foundational human rights treaty with broad human rights coverage. It is among a few human rights treaties that require states to report on the suspension or derogation of civil and political rights only under specific situations of “public emergency which threatens the life of the nation” [@iccpr]. The other treaties with this requirement include the European Convention Protection of Human Rights and Fundamental Freedoms and the American Convention on Human Rights. However, the ICCPR is the only one with global membership. As of 2023, 173 states are party to the ICCPR. This means that most of the UN's 193 members had the ability to submit derogations.

@@

@@

We expect that states that derogated with the intention to impose emergency policies will follow through. Many argue that states ratify treaties with the expectation of reputational and other benefits [@Hathaway:2007; @Hafner-BurtonLeVeckVictor:2017]. However, derogations have no such public audience and are not signaling an increased commitment to human rights. There is no clear or expected connection with foreign aid, trade, or other theorized ‘rewards’ for larger legal actions. Instead, derogations are a very specific legal action that states voluntarily enter into after commitment has fully taken place. Given this, we expect that derogating states took the legal action seriously and generally followed through with the derogations they submitted.

@@

We use a binary indicator of whether a state was under ICCPR derogation conditions during a given week. We collected and coded this data from real time state action updates recorded at the UN Treaty Collection [@UnitedNations:2023]. During the 15 months covered in this project, there were 115 formal derogations submitted mentioning 8 specific ICCPR articles; 15 derogations made no reference to any specific article. Of the derogations that indicated specific time limits, the average derogation duration was 41 days. We mark the start and end date of each derogation so that each country-week observation indicates whether a state has derogated.

>>>>

---
title: "Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19"
short-title: "Pandemic Pass?"
subtitle: "(Research note)"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  orcid: 0000-0003-3506-7590
  url: https://www.suparnachaudhry.com/
  email: schaudhry@lclark.edu
- name: Audrey L. Comstock
  affiliation: Arizona State University
  orcid: 0000-0003-1772-5350
  url: https://audreylcomstock.weebly.com/
  email: audrey.comstock@asu.edu
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  orcid: 0000-0002-3948-3914
  url: https://www.andrewheiss.com
  email: aheiss@gsu.edu
  attributes:
    corresponding: true
correspondence-prefix: "Correspondence concerning this article should be addressed to"
abstract: |
  This research note asks whether states issuing pandemic-era human rights treaty derogations implemented emergency provisions as intended or used them to abuse human rights during a time of crisis. In an effort to combat the COVID-19 pandemic, many countries declared states of emergency and derogated (temporarily suspended) from their international human rights treaty obligations. Using data from the Varieties of Democracy PanDem dataset and the Oxford COVID-19 Government Response Tracker, we find that states that derogated from their international human rights obligations imposed emergency measures that were temporary and did not violate non-derogable rights. On the other hand, states that did not derogate were more likely impose discriminatory measures, enact emergency measures without time limits and violate non-derogable rights. Our results support the role that flexibility mechanisms such as derogations play in international law and show that states are being sincere about their intentions and not, generally, using these mechanisms to cover abusive behavior.
thanks: |
  We would like to thank particpants of the 2022 Peace Science Society conference and the 2023 International Studies Association (ISA) convention for their comments and feedback.
date: May 16, 2024
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/mountainous-mackerel>"
keywords:
  - human rights
  - treaty derogations
  - COVID-19
  - emergency policies
bibliography: references.json
csl: csl/chicago-author-date.csl
link-citations: true
---

@@

The ICCPR is a foundational human rights treaty with broad human rights coverage. It is among a few human rights treaties that require states to report on the suspension or derogation of civil and political rights only under specific situations of “public emergency which threatens the life of the nation” [@iccpr]. The other treaties with this requirement include the European Convention on Human Rights (ECHR) and the American Convention on Human Rights (ACHR). Most of the states that derogated from the regional human rights conventions also derogated from the ICCPR, demonstrating that regional derogation activity did not preclude or replace ICCPR derogation activity.[^regional] The ICCPR is the only treaty with global membership---as of 2023, 173 states are party to the ICCPR. This means that most of the UN's 193 members had the ability to submit derogations. As such, we focus on ICCPR derogation activity in this research.[^future]

[^regional]: See Table A5 for a list of countries that derogated from both the ACHR/ECHR and the ICCPR.

[^future]: This project specifically examines what happens after derogations are filed. In a separate project by Authors (2024) we explore the determinants of derogations and find that (1) pandemic-era backsliding and (2) the severity of the pandemic as measured by national-level deaths are significant and positive indicators of pandemic-related derogation submission (see Table A6 in the appendix).

@@

States follow different pathways to the actual submission of ICCPR derogations based on national level norms and policies. The UN does not specify the set pathway for national derogation approval. Most states followed a pathway of executive level instruction and authorization. For example, Argentina's derogations were described, upon submission to the UN as following an emergency decree issued by its executive branch [@un-derog-arg]. Guatemala directly referenced the will of the President of the state in initiating the derogation policy [@un-derog-gtm]. Some states such as the Dominican Republic, specified the role of the legislature in advancing the derogation provision [@un-derog-dom]. Other states, such as Azerbaijan, referenced how the executive and legislature were involved in approving [@un-derog-aze]. In reading through the written derogations, eighteen of the issuing states indicated that the executive or head of state directed the derogation, and five indicated the national legislature's role in approving it. All written derogations referenced a national declaration of emergency.

Almost all states that issued derogations specified time limits, as the derogation format requires. Fifteen of the derogating states issued derogations that extended or renewed prior ones. The frequency of renewal signifies that states took the times specified within the derogations seriously and updated the treaty action when their national government expected the times of crises to continue. We understand the time specification to be a high level of precision about obligation, applying the dimensions of legalization [@AbbottKeohaneMoravcsik:2000] which contributes to a higher expectation of compliance, or adherence with the treaty.

@@

We expect that states that derogated with the intention to impose emergency policies will follow through. Many argue that states ratify treaties with the expectation of reputational and other benefits [@Hathaway:2007; @Hafner-BurtonLeVeckVictor:2017]. However, derogations have no such public audience and are not signaling an increased commitment to human rights. There is no clear or expected connection with foreign aid, trade, or other theorized 'rewards' for larger legal actions. Instead, derogations are a very specific legal action that states voluntarily enter into after commitment has fully taken place. Given this, we expect that derogating states took the legal action seriously and generally followed through with the derogations they submitted.

@@

We use a binary indicator of whether a state was under ICCPR derogation conditions during a given week. We collected and coded this data from real time state action updates recorded at the UN Treaty Collection [@UnitedNations:2023]. During the 15 months covered in this project, there were 115 formal derogations submitted mentioning 8 specific ICCPR articles; 15 derogations made no reference to any specific article. Of the derogations that indicated specific time limits, the average derogation duration was 41 days. Fifteen of the derogating states issued renewals/extensions of derogations that had time limits. We mark the start and end date of each derogation so that each country-week observation indicates whether a state has derogated.
